The Magnetorotational Instability Prefers Three Dimensions

### 2019-11-11 paper/paper.tex
edits to section

The calculation becomes very messy. We therefore only sketch the details, and report the main results.
We analyse the growth rate as a function of  $k_{y},k_{z}$ and $\SSC$ near $k_{y} \approx0$.  We therefore expand the whole problem in a power-series ansatz
\Beq
\gamma(\SSC,k_{y},k_{z}) \ = \   \sum_{n\ge 0} \gamma_{n}(\SSC,k_{z}) k_{y}^{n}. 
\Eeq
We use the same asymptotic techniques as before to obtain each progressive correction. We solve the leading order by virtue that all non-constant coefficients vanish for 2D modes. The following dispersion relations determine the maximum growth rate and the corresponding wavenumber,
\Beq
\mathfrak{D}_{0}(k_{z}) &=& \frac{f k_z^2 \left(B_0^2 S k_z^2+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_z^2+\gamma _0^2\right){}^2}+\frac{\pi ^2}{d^2}+k_z^2 \ = \ 0 \\
\frac{\mathfrak{D}_{0}'(k_{z})}{2k_{z}} &=& \frac{\gamma _0^2 f \left(B_0^2 k_{z}^2 (S-f)+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_{z}^2+\gamma _0^2\right)^3}+1 \ = \ 0.
\Eeq
We transform these relations into polynomials in terms of non-dimensional variables,
\Beq
\label{dispersion}
\y^2 \s^2 \Ro^2 +  \y \, \s  \, ( \y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s ) \x + ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x^2 +\x^3 &=& 0 \\ 
\y \, \s  \, (\y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s    )  + 2 ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x +3\x^2 &=& 0,
\label{critical-dispersion}
\Eeq
where
\Beq
\x  \ = \ \frac{d^2 k_z^2}{\pi ^2}, \quad  \quad \y \ = \ \frac{\gamma_{0}^{2}}{S^{2}}, \quad \quad \s  \ = \ \frac{S}{S_{c}}.
\Eeq
These are directly related to the parameters reported on figure~\ref{fig:growth_rate}.
The two polynomials are straightforward to solve numerically for $\x,\y$ given $\s, \Ro$. However we  still want to find critical values for $\s$, where the fastest growing modes are two dimensional. From now on, the calculation becomes sufficiently complicated that we only outline the methodology.

We find the higher-order $\gamma_{n}(\s,k_{z})$ coefficients by applying order-by-order solvability conditions. We first find that $\gamma_{1}(\s, k_{z}) = 0$ identically. This renders the first-order correction problem solvable, which means solving it and moving on the the 2nd-order corrections. The structure of the growth rate is therefore 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2} \ + \ \ldots 
\Eeq
This is the same structure as the asymptotic calculation in the previous section. Now effectively $\Upsilon \propto \gamma_{2}$. The case with dominant 2D modes corresponds to $\gamma_{2}(\s,\x)=0$. Finding this condition is a formidable calculation, which produces the degree-6 polynomial expression  
\Beq
& \y^{6} \s^6\Ro^6  \ + \
 \y^4 \s^5 \Ro^4   \left( (3 \xi  \s+6 \y-1)\Ro -3 \xi  \s \right) \, \x +
  \nonumber \\  
& \frac{ \y^3 \s^4 \Ro^2  }{4}  \left( (60 \y -4 \xi  (\s-8) \s-16 + 3 \s^2)\Ro^{2} + 2 \s (4 \xi  (\s+1)-3 \s) \Ro - (4 \xi -3 ) \s^2 \right) \, \x^{2} + \nonumber \\ 
& \frac{\y^2  \s^3 \Ro }{2}  \left( (3 \s^2-4 \xi  (\s-3) \s+40 \y-12) \Ro^{2} + 24 \xi \s \Ro + (4 \xi -3) s^2 \right) \, \x^{3} + 
\nonumber \\ 
&  \frac{\y \,\s^2}{4} \left( (60 \y -(4 \xi -3) \s^2 - 16)\Ro^{2} + 2\s(3 \s-4 \xi  (\s-3)) \Ro - (4 \xi -3) \s^2 \right)\, \x^{4} +\nonumber \\ 
 &  \s\, (( 6 \y -\xi  \s-1) \Ro- \xi  \s) \, \x^{5}  \ + \   
   \x^{6} \ = \ 0.
   \label{2D-condition}
\Eeq
The new coefficient is
\Beq
\xi \ = \ \frac{\pi^{2}-6}{12} \ = \ \frac{\int_{-d/2}^{d/2} x^2 \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}{\frac{d^{2}}{\pi^{2}}\int_{-d/2}^{d/2}  \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}.
\Eeq
This results from projecting the squared shear onto the leading-order eigenmodes, just as in the previous section.

>>>>

We analyse the growth rate as a function of  $k_{y},k_{z}$ and $\SSC$ near $k_{y} \approx0$.  We therefore expand the whole problem in a power-series ansatz
\Beq
\gamma(\SSC,k_{y},k_{z}) \ = \   \sum_{n\ge 0} \gamma_{n}(\SSC,k_{z}) k_{y}^{n}, \quad \psi(x) \ = \ \sum_{n\ge 0} \psi_{n}(x) k_{y}^{n}
\Eeq
We use the same asymptotic techniques as before to obtain each progressive correction. 
\Beq
\psi_{0}'' - \Phi_{0} \psi_{0} & = & 0 \label{eq:psi0} \\
\psi_{1}'' - \Phi_{0} \psi_{1} & = & \Phi_{1} \psi_{0} \label{eq:psi1} \\
\psi_{2}'' - \Phi_{0} \psi_{2} & = & \Phi_{2} \psi_{0} + \Phi_{1} \psi_{1} \label{eq:psi2}  
\Eeq
We solve the leading order by virtue that all non-constant coefficients vanish for 2D modes. The leading-order eigenfunction is $\psi_{0} = \cos(\pi x / d)$, just as before. The following dispersion relations determine the maximum growth rate and the corresponding wavenumber,
\Beq
\mathfrak{D}_{0}(k_{z}) &=& \frac{f k_z^2 \left(B_0^2 S k_z^2+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_z^2+\gamma _0^2\right){}^2}+\frac{\pi ^2}{d^2}+k_z^2 \ = \ 0 \\
\frac{\mathfrak{D}_{0}'(k_{z})}{2k_{z}} &=& \frac{\gamma _0^2 f \left(B_0^2 k_{z}^2 (S-f)+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_{z}^2+\gamma _0^2\right)^3}+1 \ = \ 0.
\Eeq
We transform these relations into polynomials in terms of non-dimensional variables,
\Beq
\label{dispersion}
\y^2 \s^2 \Ro^2 +  \y \, \s  \, ( \y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s ) \x + ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x^2 +\x^3 &=& 0 \\ 
\y \, \s  \, (\y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s    )  + 2 ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x +3\x^2 &=& 0,
\label{critical-dispersion}
\Eeq
where
\Beq
\x  \ = \ \frac{d^2 k_z^2}{\pi ^2}, \quad  \quad \y \ = \ \frac{\gamma_{0}^{2}}{S^{2}}, \quad \quad \s  \ = \ \frac{S}{S_{c}}.
\Eeq
These are directly related to the parameters reported on figure~\ref{fig:growth_rate}.
The two polynomials are straightforward to solve numerically for $\x,\y$ given $\s, \Ro$. However we  still want to find critical values for $\s$, where the fastest growing modes are two dimensional. From now on, the calculation becomes sufficiently complicated that we only outline the methodology.

We find the higher-order $\gamma_{n}(\s,k_{z})$ coefficients by applying order-by-order solvability conditions. We first find that $\gamma_{1}(\s, k_{z}) = 0$ identically. This renders the first-order correction \eq{eq:psi1} solvable. The 1st-order eigenfunction takes the form 
\Beq
\psi_{1}(x) \propto F(\pi x/d) \quad \text{where} \quad  F(\vartheta) = \vartheta (\vartheta \sin (\vartheta)+\cos (\vartheta))-\frac{\pi^{2}}{4} \sin (\vartheta).
\Eeq
The structure of the growth rate is therefore 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2} \ + \ \ldots 
\Eeq
This is the same structure as the asymptotic calculation in the previous section. Now effectively $\Upsilon \propto \gamma_{2}$. The case with dominant 2D modes corresponds to $\gamma_{2}(\s,\x)=0$.  The final result derives from the 2nd-order solvability condition 
\Beq
\int_{-d/2}^{d/2} ( \Phi_{2} \psi_{0} + \Phi_{1} \psi_{1} ) \psi_{0} \dd{x}  \ = \ 0,
\Eeq
which produces the degree-6 polynomial expression  
\Beq
& \y^{6} \s^6\Ro^6  \ + \
 \y^4 \s^5 \Ro^4   \left( (3 \xi  \s+6 \y-1)\Ro -3 \xi  \s \right) \, \x +
  \nonumber \\  
& \frac{ \y^3 \s^4 \Ro^2  }{4}  \left( (60 \y -4 \xi  (\s-8) \s-16 + 3 \s^2)\Ro^{2} + 2 \s (4 \xi  (\s+1)-3 \s) \Ro - (4 \xi -3 ) \s^2 \right) \, \x^{2} + \nonumber \\ 
& \frac{\y^2  \s^3 \Ro }{2}  \left( (3 \s^2-4 \xi  (\s-3) \s+40 \y-12) \Ro^{2} + 24 \xi \s \Ro + (4 \xi -3) s^2 \right) \, \x^{3} + 
\nonumber \\ 
&  \frac{\y \,\s^2}{4} \left( (60 \y -(4 \xi -3) \s^2 - 16)\Ro^{2} + 2\s(3 \s-4 \xi  (\s-3)) \Ro - (4 \xi -3) \s^2 \right)\, \x^{4} +\nonumber \\ 
 &  \s\, (( 6 \y -\xi  \s-1) \Ro- \xi  \s) \, \x^{5}  \ + \   
   \x^{6} \ = \ 0.
   \label{2D-condition}
\Eeq
The new coefficient is
\Beq
\xi \ = \ \frac{\pi^{2}-6}{12} \ = \ \frac{\int_{-d/2}^{d/2} x^2 \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}{\frac{d^{2}}{\pi^{2}}\int_{-d/2}^{d/2}  \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}.
\Eeq
This results from projecting the squared shear onto the leading-order eigenmodes, just as in the previous section.
